The fraction of polar aligned circumbinary disks

### 2025-04-22 src/tex/response.tex

\newenvironment{response}[1]{
    \fcolorbox{blue}{white}{
    \begin{minipage}{0.46\textwidth}
        #1
    \end{minipage}
    }\hfill
    \begin{minipage}{0.46\textwidth}
    }
    {\end{minipage}}

\begin{document}

\begin{letter}{}
    \opening{Dear Editor Carraro, Data Editors, and Anonymous Reviewer,}

    Please find our revised manuscript alongside the below response to the reviewer's report. The authors would like to thank the editor, reviewer, and data editors for their thoughtful review, which has lead to a greatly improved manuscript. Changes have been highlighted in \textcolor{blue}{\bf blue}.

    In addition to responding to the reviewer's report, the authors have added text referencing the recent strong evidence for the first polar circumbinary planet\footnote{
    Baycroft, T. A., Sairam, L., Triaud, A. H. M. J., \& Correia, 760
A. C. M. 2025, Science Advances, 11, eadu0627,761 doi: \href{http://doi.org/10.1126/sciadv.adu0627}{10.1126/sciadv.adu0627}}. Please see lines 41-46 and 711-714.

@@

    \begin{response}
        {Fig. 8: The bottom part of this figure could be further commented in Sect. 6.1.}
        We have added text in Section 6.1 discussing this panel.
    \end{response}

    \begin{response}{
        L532: Bohn+2022 could be a relevant reference to add here.
    }
     Thank you for the great suggestion. We have added the reference.   
    \end{response}

    \begin{response}
        {L580: I would conclude the discussion (or add to the conclusion) that, besides imposing constraints on the disc radial extend and j, the two studied mechanisms (disc breaking and self-gravity) are expected to lower the fraction of polar CBDs. If I understood correctly, large values of j and radially extended discs are unlikely to become polar (as a whole at least). Therefore, the statistical relevance/weight of this subgroup of discs should be less important at the overall polar CBD population level. This could be briefly commented at the end.}

        This is a great suggestion, however the overall effect is more complicated. The main effect of both mechanisms is to shorten the alignment timescale so that fewer disks are left in an arbitrarily misaligned state at the end of their lives. We have added text explaining how this process works for gravitationally unstable disks and commenting on the shortened timescale for both effects.
    \end{response}

@@

    \begin{response}{
        L628: Sections (missing s).
    }
    This has been fixed.
    \end{response}

    \begin{response}{
        L642: At the end of the conclusions (or else at the end of the discussion), a brief discussion on how future observations could test these findings would add value to the article.
    }
        We have added a discussion about the possibility of confirming this work though observations. We propose a survey to measure the inclinations of disks in young eclipsing binary systems could reasonably constrain the polar fraction.
    \end{response}

    \begin{response}{
        Data Editor's Comment
    }
    Author's Response
    \end{response}

    \begin{response}
        {I. Is this a showyourwork paper? If it is then we might be missing files necessary to properly augment the final article with code links. And we would need to add a paragraph to the text explaining the showyourwork icons. Would you contact the Data Editors helpdesk if you need assistance ensuring that the showyourwork elements show up in the final article. }
        Yes, this is a \texttt{showyourwork} paper. We have added the necessary files to generate code links and added a sentence that explains the icons.
    \end{response}

    \begin{response}
        {
            II. I don't see the Zenodo DOI results tagged in the text. Can that be added using the instructions here: \url{https://journals.aas.org/aastexguide/\#datasets}
        }
        We have added the Zenodo DOI for this dataset to the caption of Figure 3.
    \end{response}

    \begin{response}
        {
            III. For Showyourwork papers, we would like to you to freeze the final final version on GitHub and add that Zenodo DOI to the text as well. More information is here: \url{https://journals.aas.org/aastexguide/\#datasets}

            [2] \url{https://journals.aas.org/data-guide/\#repositories}

[3] \url{https://docs.github.com/en/repositories/archiving-a-github-repository/referencing-and-citing-content} 
        }

        We have frozen the final version of the repository and placed the Zenodo DOI in the text.
    \end{response}

@@

    \begin{response}
        {
            IV. It appears you intend tedjohnson12/polar-disk-freq to be reusable software. If so we ask that you again freeze a copy of the final final version and cite that frozen copy with the final article.

[1] \url{https://journals.aas.org/policy-statement-on-software/}

[2] \url{https://journals.aas.org/data-guide/\#repositories}

### 2025-04-22 src/tex/ms.tex
Sync changes from Rebecca + Steve Lubow

Observed circumbinary planets -- those orbiting two stars -- have mostly been found to orbit in the same plane as their hosts \citep{doyle2011,orosz2012,welsh2012}. While this preference is the result of target selection effects (\citealt{martin2014}, see also \citealt{martin2015}, \citealt{martin2017a}), it is also representative of the fact that, for the disk progenitors of these planets \citep[from which the planets inherit orbital parameters, see][]{childs2021}, a coplanar alignment is one of two possible stable orientations \citep{bate2000,lubow2000}. The other configuration -- a polar disk -- is characterized by being inclined $90^\circ$ relative to the binary orbital plane \citep{aly2015,martin2017}. \RR{O}nly a handful of confirmed polar disks have ever been discovered. These include a polar debris disk in the 99 Her system (\citealt{kennedy2012}, which likely evolved from a polar gas disk, see \citealt{smallwood2020}) and polar gas disks in the HD 98800 \citep{kennedy2019} and V773 Tau \citep{kenworthy2022} systems. \RR{Currently the only evidence for a polar circumbinary planet comes from AC Her, a post-AGB star system that hosts an unusually-truncated circumbinary disk, with a planet located interior to the disk's inner edge being the most likely cause \citep{hillen2015,anugu2023,martin2023}.}

@@

Given enough time, a gas disk that is misaligned from one of the two stationary inclinations moves to alignment through torques excited by warps that form as the disk undergoes differential precession \citep{bate2000,lubow2000}. Disks which are misaligned from coplanar only slightly (including the anti-aligned case where the mutual inclination $i\sim\pi$) undergo prograde or retrograde nodal precession \citep{papaloizou1995}. \citet{bate2000} and \citet{lubow2000} show that this kind of motion leads to coplanar alignment on a timescale that scales as $h^2/\alpha$, where $h$ and $\alpha$ are the disk aspect ratio and viscosity parameter, respectively.

@@

For an isotropic angular momentum distribution \RR{(i.e. one where the specific angular momentum $\bm{\ell}$ has no preference for any direction)}, the polar fraction in this limit is
\begin{align}
    f_{{\rm polar}, j \gg 1} &= \int_0^\pi\,\left (1-\frac{2}{\pi}\arcsin{\left (\sqrt{\frac{2}{5}} \csc{i} \right )}\right ) \sin{i}\,di\, , \\
    & = \variable{output/high_j_integral.txt} \, . \nonumber
\end{align}
This is the value that the higher $j$ curves in Figure \ref{fig:ebin} tend to for low $e_{\rm b}$, computed numerically.

@@

\begin{figure*}
    \begin{centering}
        \includegraphics[width=\textwidth]{figures/compare_ebin_idist.pdf}
        \caption{
            Same as Figure \ref{fig:ebin}, but with the mutual inclination distribution given by equation (\ref{eq:idist}). \RR{  The insert shows the mutual inclination probability distribution normalized by $\sin i$. This normalization shows that the probability-per-unit-area on the unit sphere becomes very high at $i=0$ (coplanar alignment) when $\sigma$ is small (dashed and dotted lines), but it is uniform everywhere in the isotropic case ($\sigma \rightarrow \infty$, solid line).   For low $j$, lower inclination preferences decrease $f_\text{polar}$ around $e_\text{b}\sim0.5$, but have little effect near $e_{\rm b}=0$ and 1. For high-$j$, low inclinations lower the value that $f_\text{polar}$ tends to as $e_\text{b}\rightarrow 0$.}
        }
        \script{compare_ebin_idist.py}
        \label{fig:ebin-idist}
    \end{centering}
\end{figure*}

@@

The maximum value of $j$ to be stable against self-gravity is
\begin{multline}
    \label{eq:jmax-shallow}
    j_\text{max} = \frac{2h}{f (1-f) (5/2 - p)\sqrt{1-e_\text{b}^2}}\\ \sqrt{\frac{r_\text{in}}{a_\text{b}}} \left(\frac{r\text{out}}{r_\text{in}}\right)^{p-2} \left( \left( \frac{r_\text{out}}{r_\text{in}} \right)^{5/2 - p}- 1\right)\, .
\end{multline}

@@

\RR{
    When disk breaking and gravitational instability are considered jointly it may be that $f_\text{polar}$, as calculated in this study, is not as relevant as ``the fraction of binary systems which host a polar circumbinary disk.'' Gravitational instability (if not suppressed by heating as mentioned previously) requires that massive high-$j$ disks be more radially extended, making them more susceptible to differential precession and breaking. The result is that a circumbinary system initially hosting a single, extended, misaligned disk may evolve to host a series narrower rings, with the innermost most likely to have completed its evolution to a coplanar or polar alignment.
}

@@

\RR{
These results for $f_\text{polar}$ can be tested through population-level observations of circumbinary disk systems. In practice, however, the mutual inclination between the disk and its binary host can be difficult to constrain, and therefore polar disks can be difficult to distinguish from coplanar. Known misaligned disks rely on fortunate geometry to constrain the mutual inclination, such as an eclipse of a nearby binary in the case of V773 Tau B \citep{kenworthy2022}. However, $f_\text{polar}$ could be statistically constrained by looking for polar disks around young eclipsing binaries because the binary-observer inclination is known. Imaging of a face-on disk in such systems would confirm a polar configuration. However, for an edge-on disk it would be difficult to differentiate between a coplanar and polar due to a degeneracy in viewing angle. A carefully done survey of a large population of eclipsing binaries \citep[e.g., those found by the Kepler Mission][]{kirk2016} would likely constrain the fraction of polar circumbinary disks.}

\RR{
A similar method has been proposed by \citet{zhang2019} to distinguish polar from coplanar planets in eclipsing binary systems via eclipse timing variations. I contrast to the imaging method suggested previously, such a method would be less expensive, have fewer degeneracies, and have many more potential targets as it would not be restricted to young systems. Additionally, it would be most sensitive to planets with small separations from their host binaries, meaning the alignment timescale for the progenitor disk would be much shorter than for those that could be resolved via imaging. Together, these two methods could provide robust constraints of $f_\text{polar}$ across many regimes of host parameters and disk separation.
}

>>>>

Observed circumbinary planets -- those orbiting two stars -- have mostly been found to orbit in the same plane as their hosts \citep{doyle2011,orosz2012,welsh2012}. While this preference is the result of target selection effects (\citealt{martin2014}, see also \citealt{martin2015}, \citealt{martin2017a}), it is also representative of the fact that, for the disk progenitors of these planets \citep[from which the planets inherit orbital parameters, see][]{childs2021}, a coplanar alignment is one of two possible stable orientations \citep{bate2000,lubow2000}. The other configuration -- a polar disk -- is characterized by being inclined $90^\circ$ relative to the binary orbital plane \citep{aly2015,martin2017}. \RR{O}nly a handful of confirmed polar disks have ever been discovered. These include a polar debris disk in the 99 Her system (\citealt{kennedy2012}, which likely evolved from a polar gas disk, see \citealt{smallwood2020}) and polar gas disks in the HD 98800 \citep{kennedy2019} and V773 Tau \citep{kenworthy2022} systems.

\RR{Recently, strong evidence for the first polar planet has been presented by \citet{baycroft2025}. They use radial velocity measurements of an eclipsing binary to infer retrograde apsidal precession which can be best explained by a planetary mass companion in a polar orbit.
The only previous evidence for a polar circumbinary planet comes from AC Her, a post-AGB star system that hosts an unusually-truncated circumbinary disk, with a planet located interior to the disk's inner edge being the most likely cause \citep{hillen2015,anugu2023,martin2023}.
}

@@

Given enough time, a gas disk that is misaligned from one of the two stationary inclinations moves to alignment through torques excited by warps that form as the disk undergoes differential precession \citep{bate2000,lubow2000}. Disks which are misaligned from coplanar only slightly \RR{(or slightly misaligned from retrograde coplanar $i\sim\pi$) undergo  retrograde (or prograde)} nodal precession \citep{papaloizou1995}. \citet{bate2000} and \citet{lubow2000} show that this kind of motion leads to coplanar alignment on a timescale that scales as $h^2/\alpha$, where $h$ and $\alpha$ are the disk aspect ratio and viscosity parameter, respectively.

@@

For an isotropic \RR{distribution of the angular momentum vector direction (i.e. one where the specific angular momentum $\bm{\ell}$ has no preference for any direction)}, the polar fraction in this limit is
\begin{align}
    f_{{\rm polar}, j \gg 1} &= \int_0^\pi\,\left (1-\frac{2}{\pi}\arcsin{\left (\sqrt{\frac{2}{5}} \csc{i} \right )}\right ) \sin{i}\,di\, , \\
    & = \variable{output/high_j_integral.txt} \, . \nonumber
\end{align}
This is the value that the higher $j$ curves in Figure \ref{fig:ebin} tend to for low $e_{\rm b}$, computed numerically.

@@

\begin{figure*}
    \begin{centering}
        \includegraphics[width=\textwidth]{figures/compare_ebin_idist.pdf}
        \caption{
            Same as Figure \ref{fig:ebin}, but with the mutual inclination distribution given by equation (\ref{eq:idist}) \RR{that is plotted in the inset normalized by $\sin i$. This normalization shows that the probability-per-unit-area on the unit sphere becomes very high at $i=0$ (coplanar alignment) when $\sigma$ is small (dashed ($\sigma=\pi/2$) and dotted ($\sigma=\pi/8$) lines), but it is uniform everywhere in the isotropic case ($\sigma \rightarrow \infty$, solid line).   The main figure shows that for low $j$, low $\sigma$ (low inclination preference) decreases $f_\text{polar}$. For high-$j$, low $\sigma$ (low inclination preferences) lowers the value that $f_\text{polar}$ tends to as $e_\text{b}\rightarrow 0$.}
        }
        \script{compare_ebin_idist.py}
        \label{fig:ebin-idist}
    \end{centering}
\end{figure*}

@@

The maximum value of $j$ to be stable against self-gravity is
\begin{multline}
    \label{eq:jmax-shallow}
    j_\text{max} = \frac{2h}{f (1-f) (5/2 - p)\sqrt{1-e_\text{b}^2}}\\ \sqrt{\frac{r_\text{in}}{a_\text{b}}} \left(\frac{r_\text{out}}{r_\text{in}}\right)^{p-2} \left( \left( \frac{r_\text{out}}{r_\text{in}} \right)^{5/2 - p}- 1\right)\, .
\end{multline}

@@

\RR{
    When a disk does become gravitationally unstable, fragments form and heat up. If the cooling time is greater than a few times the orbital period then fragments collide before they can collapse, leading to a turbulent steady-state with an enhanced effective $\alpha$ viscosity \citep{gammie2001}. Increasing $\alpha$ shortens the evolutionary timescale, so, just like disk breaking, gravitational instability in this regime can result in faster alignment and more disks that reach their polar or coplanar steady-state.
}

@@

\RR{
These results for $f_\text{polar}$ can be tested through population-level observations of circumbinary disk systems. In practice, however, the mutual inclination between the disk and its binary host can be difficult to constrain, and therefore polar disks can be difficult to distinguish from coplanar. Known misaligned disks rely on fortunate geometry to constrain the mutual inclination, such as an eclipse of a nearby binary in the case of V773 Tau B \citep{kenworthy2022}. However, $f_\text{polar}$ could be statistically constrained by looking for polar disks around young eclipsing binaries because the binary-observer inclination is known. Imaging of a face-on disk in such systems would confirm a polar configuration. However, for an edge-on disk it would be difficult to differentiate between a coplanar and polar due to a degeneracy in viewing angle. A carefully done survey of a large population of eclipsing binaries \citep[e.g., those found by the Kepler Mission,][]{kirk2016} would likely constrain the fraction of polar circumbinary disks.}

\RR{
$f_\text{polar}$ could also be inferred from the frequency of polar planets. \citet{zhang2019} show that retrograde apsidal precession of a binary can be caused by a polar circumbinary planet and develop a method to detect such precession in eclipsing binaries via eclipse timing variations. \citet{baycroft2025} successfully detect retrograde apsidal precession using radial velocity measurements, and conclude that a polar planet is the only suitable explanation. Both of these methods could be applied to the population of known eclipsing binaries. Compared to the imaging method suggested previously, planet-focused methods would be most sensitive to small separations from the binary, meaning the alignment timescale for the progenitor disk would be much shorter than for those can can be resolved via imaging. Together, however, these two methods (disk imaging, apsidal precession inference) could provide robust constraints of $f_\text{polar}$ across many regimes of host parameters and disk separation.
}
